Flat histogram method comparison on 2D Ising Model

### 2019-07-11 papers/histogram/ising.tex
fixed some ising paper errors, made the ground state part of the data plot consideration, and edited the run script to account for it.

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte-Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact} make the
system sufficiently interesting for such theoretical comparisons. It is
also of important note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the infinite volume limit ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over all enumerated spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
where $g(E)$ is the multiplicity of the system which is identical to
$D(E)$. We can estimate the average deviation from the exact solution
using the relation~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control}:
\begin{align}
\braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
\end{align}
As per the implementation by Wang and
Landau~\cite{wang2001determining}, the total number of available energy
states are $N-1$.
\jpsays{Also the necessary condition that the ground state is
$2$ for the 2D Ising model is enforced to guarantee the accuracy of the
density of states at low temperatures. If $2^N$ states is the condition
used for the ferromagnetic ising model, the rescaled factor is
determined primarily by the maximum density of states impacing the
accuracy of the energy levels at or near the ground state.}
%\section{Discrete Spin Systems} In this work, we compare the
%convergence properties of the family of Weight-Based Broad Histogram
%Methods applied to the Q-State Potts model and Ising model. The two
%classical spin systems are ideal testbeds for comparing the convergence
%of Monte-Carlo methods. The Potts and Ising models have 1st and 2nd
%order phase transitions respectively and will both require energy
%binning due to the discrete energy levels of the systems. Convergence
%scaling for the family of Weight-Based Broad Histogram methods can
%easily be explored on the two non-continuous systems.

@@

\subsection{The 32 $\times$ 32 Ising model}
For this simulation, we chose a minimum reduced temperature of $1$, which
corresponds to an interesting energy range from $-2048$ to $50$.  The number of
important energy states for this system is $N_S = 524$. \jpsays{The entropy of
this system is shown in Fig.~\ref{fig:entropy-cartoon} above, which shows that
over this energy range the entropy differs by 198, corresponding to a ratio of
$10^{86}$ between the highest and lowest density of states.}

@@

\subsection{128 $\times$ 128 Ising system}
For this simulation, we chose a minimum reduced temperature of $1$, which
corresponds to an interesting energy range from $-30268$ to $100$.  The number of
important energy states for this system is $N_S = 524$.

Fig.~\ref{fig:n128} shows the average error in the entropy as a function of time
for this system with the displacement distance of $0 = 0.05$. The solid/dashed
lines represent the average of the absolute value of the error in the entropy
averaged over eight simulations using different random number seeds. The range
of average errors for each simulation is shown as a shaded region around its
mean error. By the time $10^12$ moves have been made all but the WL and SAD
simulations have begun to converge as $1/\sqrt{t}$. We then see the WL error
saturate around $10^12$ moves. The error in the SAD method will continue to
converge faster than $1/\sqrt{t}$ until the update factor approaches that of
1/t-WL at which point both methods will converge equivalently.

>>>>

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte-Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact} make the
system sufficiently interesting for such theoretical comparisons. It is
also of important note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the infinite volume limit ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over all enumerated spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
where $g(E)$ is the multiplicity of the system which is identical to
$D(E)$. We can estimate the average deviation from the exact solution
using the relation~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control}:
\begin{align}
\braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
\end{align}
As per the implementation by Wang and
Landau~\cite{wang2001determining}, the total number of available energy
states are $N-1$.
% \jpsays{Also the necessary condition that the ground state is
% $2$ for the 2D Ising model is enforced to guarantee the accuracy of the
% density of states at low temperatures. If $2^N$ states is the condition
% used for the ferromagnetic ising model, the rescaled factor is
% determined primarily by the maximum density of states impacing the
% accuracy of the energy levels at or near the ground state.}
%\section{Discrete Spin Systems} In this work, we compare the
%convergence properties of the family of Weight-Based Broad Histogram
%Methods applied to the Q-State Potts model and Ising model. The two
%classical spin systems are ideal testbeds for comparing the convergence
%of Monte-Carlo methods. The Potts and Ising models have 1st and 2nd
%order phase transitions respectively and will both require energy
%binning due to the discrete energy levels of the systems. Convergence
%scaling for the family of Weight-Based Broad Histogram methods can
%easily be explored on the two non-continuous systems.

@@

\subsection{The 32 $\times$ 32 Ising model}
For this simulation, we chose a minimum reduced temperature of $1$, which
corresponds to an interesting energy range from $-2048$ to $48$.  The number of
important energy states for this system is $N_S = 529$.

@@

\subsection{128 $\times$ 128 Ising system}
For this simulation, we chose a minimum reduced temperature of $1$, which
corresponds to an interesting energy range from $-32768$ to $100$.  The number of
important energy states for this system is $N_S = 8214$.
Fig.~\ref{fig:n128} shows the average error in the entropy as a function of time
for this system with the displacement distance of $0 = 0.05$. The solid/dashed
lines represent the average of the absolute value of the error in the entropy
averaged over eight simulations using different random number seeds. The range
of average errors for each simulation is shown as a shaded region around its
mean error. By the time $10^{12}$ moves have been made all but the WL and SAD
simulations have begun to converge as $1/\sqrt{t}$. We then see the WL error
saturate around $10^{12}$ moves. The error in the SAD method will continue to
converge faster than $1/\sqrt{t}$ until the update factor approaches that of
1/t-WL at which point both methods will converge equivalently.
